---
title: "Global Patterns of Rabies Mortality"
format:
  dashboard:
    expandable: false
    nav-buttons: [github]
    github: https://github.com/pmmagusiak/rabies_ds_quarto_shiny_dashboard_project
    logo: images/logo.png
    theme: 
      - sandstone
      - style/custom.scss
server: shiny
include-before-body:
  text: |
    <script>
      $(document).ready(function() {
        var heights= $('.value-box-title').map(function() {
          return $(this).height();
        }).get();
        $('.value-box-title').css('height', Math.max(...heights));
      })
    </script>
date: 2025-01-23
---

```{r}
#| label: load-packages
#| context: setup
#| message: false

library(tidyverse)
library(data.table)
library(skimr)
library(janitor)
library(countrycode)
library(DT)
library(shiny)
library(bslib)
library(bsicons)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(shinyWidgets)
library(plotly)
```

```{r}
#| label: load-data
#| context: setup
#| message: false

# Importing files into R environment

human_rabies_death_raw <- fread("data/reported_number_of_human_rabies_death.csv")

hdi_raw <- fread("data/human_development_index.csv")
```

```{r}
#| label: initial-prep-data
#| context: setup
#| include: false
#| message: false

## Focusing on human_rabies_death_raw database

# human_rabies_death_raw |> filter(if_any(everything(), is.null)) # Checking if there are any rows with null values

# skim_without_charts(human_rabies_death_raw) # Using skimr package to summary data tables and overview missing values, duplicates and other information, 

# glimpse(human_rabies_death_raw) # Assessing columns which are useful for my analysis

human_rabies_death_processed <- human_rabies_death_raw |>
  filter(Period != 2022) |> 
  select(
    ParentLocationCode,
    ParentLocation,
    SpatialDimValueCode,
    Location,
    Period,
    FactValueNumeric
  ) |>
  clean_names() |> # janitor package to help tidying.
  mutate(location = countrycode(spatial_dim_value_code, "iso3c", "country.name")) 

## Focusing on hdi_raw database, initial procedures the same as above

# hdi_raw |> filter(if_any(everything(), is.null))

# skim_without_charts(hdi_raw)

# glimpse(hdi_raw) # This database has multiple columns containing country development indicators per year, it is more suitable to pivot them

hdi_processed <- hdi_raw |>
  pivot_longer(
    cols = c(-(1:5), -166, -581),
    names_pattern = "^(.*)(\\(\\d+\\))$",
    names_to = c(".value", "year")
  ) |>
  mutate(year = parse_number(year)) |>
  filter(between(year, 2010, 2021)) |> 
  select(1:5, 8, 9, 15, 20) |>
  clean_names() |> 
  mutate(country = countrycode(iso3, "iso3c", "country.name")) # I prefer to leave many variables for future analysis, for current dashboard I will only need HDI columns

## Joining the databases

# human_rabies_death_processed |> count(n_distinct(location)) # this db has 194 different countries

# hdi_processed |> count(n_distinct(country)) # this db has 195 different countries

# human_rabies_death_processed |>
#   anti_join(hdi_processed, join_by(spatial_dim_value_code == iso3)) |>   
#   count(spatial_dim_value_code) # Finding unmatched records, here and below

# hdi_processed |>
#   anti_join(human_rabies_death_processed, join_by(iso3 == spatial_dim_value_code)) |> count(iso3)

human_rabies_death_final <- human_rabies_death_processed |>
  filter(!spatial_dim_value_code %in% c("COK", "NIU"))

hdi_final <- hdi_processed |>
  filter(!iso3 %in% c("HKG", "LIE", "PSE")) |> 
  filter(!human_development_groups == "") |> # Filtering out Somalia, Monaco, Nauru and North Korea
  mutate(human_development_groups = factor(
    human_development_groups,
    levels = c("Low", "Medium", "High", "Very High")
  ))

hdi_human_rabies_death_joined <- hdi_final |>
  left_join(
    human_rabies_death_final,
    join_by(country == location, year == period)
  ) |>
  rename(number_of_deaths = fact_value_numeric) |> 
  select(-undp_developing_regions, -spatial_dim_value_code) |>
  select(1:2, 9:10, 5:8, 4, 3, 11)
```

```{r}
#| label: country-shapes-prep-data
#| context: setup
#| include: false
#| message: false

# Creating world map df with countries polygons, Antarctica omitted for better viewing experience. Columns not needed removed. Polishing additional info for countries.

world_map_df <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(!name %in% c("Antarctica")) |>
  select(name, iso_a3_eh, note_brk, geometry) |>
  mutate(note_brk = case_when(is.na(note_brk) ~ "", TRUE ~ note_brk)) |>
  mutate(note_brk = str_replace(
    note_brk,
    "^Self admin\\.\\; Claimed by (.+)$",
    "Self administrated, claimed by \\1"
  ))

# Correcting country names and filtering out countries without country code and Australia's administrative units. Populating world map df with rabies death data, filtering out dependent territories, creating new categorical variable for better visualization

world_map_rabies_df <- world_map_df |>
  filter(!name %in% c("Indian Ocean Ter.", "Ashmore and Cartier Is.")) |>
  filter(iso_a3_eh != -99) |>
  mutate(name = countrycode(iso_a3_eh, "iso3c", "country.name")) |>
  right_join(hdi_human_rabies_death_joined, join_by(iso_a3_eh == iso3, name == country)) |>
  mutate(number_of_deaths_categories = case_when(
    number_of_deaths == 0 ~ "0",
    number_of_deaths %in% 1:10 ~ "1-10",
    number_of_deaths %in% 11:100 ~ "11-100",
    number_of_deaths %in% 101:500 ~ "101-500",
    number_of_deaths > 500 ~ "500+",
    is.na(number_of_deaths) ~ "Data unavailable"
  )) |>
  mutate(number_of_deaths_categories = factor(number_of_deaths_categories, ordered = TRUE, levels = c("500+", "101-500", "11-100", "1-10", "0", "Data unavailable")))
```

```{r}
#| label: plot-prep-data
#| context: setup
#| include: false
#| message: false

hdi_human_rabies_death_joined |>
  group_by(country) |>
  summarise(sum_of_deaths = sum(number_of_deaths, na.rm = TRUE)) |>
  arrange(desc(sum_of_deaths)) # There is 6 countries that stands out in terms of rabies deaths

world_deaths <- hdi_human_rabies_death_joined |>
  group_by(year) |>
  summarize(global_yearly_deaths = sum(number_of_deaths, na.rm = TRUE)) |>
  mutate(label = "Global Yearly Deaths") |>
  mutate(label = factor(label)) # I created this variable to show Global Yearly Deaths timeline on plot

hdi_human_rabies_death_joined_labelled <- hdi_human_rabies_death_joined |>
  group_by(country) |>
  mutate(sum_of_deaths = sum(number_of_deaths, na.rm = TRUE)) |>
  ungroup() |>
  mutate(label = paste0(country, " (", sum_of_deaths, ")")) |>
  mutate(label = factor(label)) |>
  mutate(label = fct_reorder(label, sum_of_deaths, .desc = TRUE))
```

# Human

## Sidebar - Human {.sidebar}

::: {.sidebar-div}

```{r}
sliderTextInput(
  inputId = "year",
  label = "Choose a year:",
  choices = c(2010:2021),
  selected = 2021,
  grid = TRUE,
  force_edges = TRUE,
  width = "98%"
)
```

::: flex-inline
```{r}
actionButton("minus", label = "-", class = "btn-sm")
```

```{r}
actionButton("plus", label = "+", class = "btn-sm")
```
:::

```{r}
hdg_choices <- c("Low", "Medium", "High", "Very High")

virtualSelectInput(
  inputId = "hdg",
  label = "Choose Human Development Index group:",
  choices = hdg_choices,
  selected = hdg_choices,
  multiple = TRUE,
  width = "100%",
  dropboxWrapper = "body"
)
```

::: {.callout-div}
::: {.callout-note collapse="true"}
## Data Sources & Limitations

This dashboard presents data on reported human rabies deaths sourced from the [WHO Global Health Observatory](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/reported-number-of-human-rabies-deaths) and Human Development Index (HDI) rankings derived from the [Kaggle HDI dataset](https://www.kaggle.com/datasets/iamsouravbanerjee/human-development-index-dataset), based on the 2021 UN Human Development Report.

**Note:** Rabies death data is severely underreported, particularly in many African countries, due to limited surveillance and reporting systems. As such, the true burden of disease is likely underestimated.

*Territories shown in light grey on the map are either not listed in the WHO Global Health Observatory (mostly dependent territories) or were not ranked in the Human Development Index as of 2021.*
:::
:::

:::

## Main body

### Value boxes {height="19%"}

```{r}
value_box(
  title = "Year",
  value = textOutput("year"),
  theme = value_box_theme(bg = "#3C5B6F")
)
```

```{r}
value_box(
  title = "Total deaths",
  value = textOutput("total"),
  theme = value_box_theme(bg = "#3C5B6F")
)
```

```{r}
value_box(
  title = "Top country by human rabies deaths",
  value = textOutput("country"),
  theme = value_box_theme(bg = "#3C5B6F")
)
```

```{r}
value_box(
  title = "Top country death count",
  value = textOutput("death_count"),
  theme = value_box_theme(bg = "#3C5B6F")
)
```

```{r}
value_box(
  title = "Top country contribution to global deaths",
  value = textOutput("percent"),
  theme = value_box_theme(bg = "#3C5B6F")
)
```

### Row - Tabset of map / plot / data {.tabset height="81%"}

```{r}
#| title: Global burden of human rabies deaths

leafletOutput("map")
```

```{r}
#| title: Rabies mortality timeline – Plot

plotlyOutput("plot")

# bslib::card(full_screen = TRUE, card_body(plotlyOutput("plot")))  <- This line of code enables expand button
```

```{r}
#| title: Human rabies deaths data

rowCallback <- c(
  "function(row, data){",
  "  for(var i=0; i<data.length; i++){",
  "    if(data[i] === null){",
  "      $('td:eq('+i+')', row).html('NA')",
  "        .css({'color': 'rgb(151,151,151)', 'font-style': 'italic'});",
  "    }",
  "  }",
  "}"
)

hdi_human_rabies_death_joined_labelled |>
  select(-13) |> 
  datatable(colnames = c("ISO 3 country code", "Country", "WHO region code", "WHO region", "Year", "Human Development Index (HDI)", "HDI female", "HDI male", "HDI rank (2021)", "HDI group (2021)", "Death count", "Total deaths"), options = list(rowCallback = JS(rowCallback)))
```

```{r}
#| context: server

# Slider buttons

observeEvent(input$plus, {
  new_value <- input$year + 1
  if (new_value <= 2021) {
    updateSliderTextInput(session, "year", selected = new_value)
  }
})

observeEvent(input$minus, {
  new_value <- input$year - 1
  if (new_value >= 2010) {
    updateSliderTextInput(session, "year", selected = new_value)
  }
})

# Value boxes code

year <- reactive({
  hdi_human_rabies_death_joined |>
    filter(year == input$year) |>
    slice_head(n = 1) |>
    pull(year)
})

total_deaths <- reactive({
  hdi_human_rabies_death_joined |>
    filter(year == input$year) |>
    summarise(total_deaths = sum(number_of_deaths, na.rm = TRUE)) |>
    pull(total_deaths)
})

country <- reactive({
  hdi_human_rabies_death_joined |>
    filter(year == input$year) |>
    arrange(desc(number_of_deaths)) |>
    slice_head(n = 1) |>
    pull(country)
})

death_count <- reactive({
  hdi_human_rabies_death_joined |>
    filter(year == input$year) |>
    arrange(desc(number_of_deaths)) |>
    slice_head(n = 1) |>
    pull(number_of_deaths)
})

percent_all_deaths <- reactive({
  hdi_human_rabies_death_joined |>
    filter(year == input$year) |>
    mutate(percent_all_deaths = round(number_of_deaths / sum(number_of_deaths, na.rm = TRUE) * 100)) |>
    arrange(desc(number_of_deaths)) |>
    slice_head(n = 1) |>
    pull(percent_all_deaths)
})

output$year <- renderText({
  year()
})

output$total <- renderText({
  total_deaths()
})

output$country <- renderText({
  country()
})

output$death_count <- renderText({
  death_count()
})

output$percent <- renderText({
  paste0(percent_all_deaths(), "%")
})

# Map code

map_filtered <- reactive({
  subset(world_map_rabies_df, human_development_groups %in% input$hdg) |>
    filter(year == input$year)
})

output$map <- renderLeaflet({
  world_map_df |>
    leaflet() |>
    addProviderTiles("CartoDB.Positron", options = providerTileOptions(minZoom = 2)) |>
    addPolygons(
      color = "black",
      weight = .5,
      opacity = 1,
      label = ~ lapply(paste0("<b>", name, "</b>", "<br/>", note_brk), htmltools::HTML)
    ) |>
    setView(0, 20, zoom = 2) |>
    setMaxBounds(
      lng1 = -180, lat1 = -60,
      lng2 = 180,  lat2 = 85
    )
})

observe({
  pal <- colorFactor(palette = c("#FFF4D9", "#FFEB9A", "#FFC94A", "#F26B30", "#960018"), levels = c("0", "1-10", "11-100", "101-500", "500+"))

  leafletProxy("map", data = map_filtered()) |>
    clearGroup("dynamic") |>
    clearControls() |>
    addPolygons(
      group = "dynamic",
      color = "black",
      fillColor = ~ pal(number_of_deaths_categories),
      weight = .5,
      opacity = 1,
      fillOpacity = .7,
      highlightOptions = highlightOptions(
        weight = 2.5,
        color = "#666",
        fillOpacity = .7,
        bringToFront = TRUE
      ),
      label = ~ lapply(paste0("<b>", name, "</b>", "<br/>", "Death count: ", number_of_deaths), htmltools::HTML)
    ) |>
    addLegend("bottomright", pal = pal, values = ~number_of_deaths_categories, title = "Number of deaths", opacity = 1) |>
    setView(0, 20, zoom = 2)
})

# Plot code

output$plot <- renderPlotly({
  ggplot <- hdi_human_rabies_death_joined_labelled |>
    filter(country %in% c("China", "India", "Philippines", "Myanmar (Burma)", "Kenya", "Vietnam")) |>
    ggplot(aes(x = year, y = number_of_deaths, color = label)) +
    geom_line(linewidth = 1) +
    geom_line(
      data = world_deaths,
      aes(x = year, y = global_yearly_deaths, color = label),
      inherit.aes = FALSE,
      linewidth = 1.2,
      linetype = 3
    ) +
    labs(
      x = "Year",
      y = "Number of deaths",
      color = "Country (Total Deaths)",
      title = "Annual Human Rabies Deaths in Top-6 Highest-Burden Countries and Worldwide",
    ) +
    scale_color_manual(values = c(
      "Global Yearly Deaths" = "black", "China (9354)" = "#1f77b4",
      "India (4351)" = "#ff7f0e",
      "Philippines (3014)" = "#2ca02c",
      "Myanmar (Burma) (1318)" = "#d62728",
      "Kenya (1115)" = "#9467bd",
      "Vietnam (1038)" = "#8c564b"
    )) +
    scale_x_continuous(n.breaks = 10)

  ggplotly(ggplot) |>
    layout(xaxis = list(fixedrange = TRUE), yaxis = list(fixedrange = TRUE)) |>
    config(modeBarButtonsToRemove = c("zoomIn2d", "zoomOut2d"))
})
```

# Animal
